---
title: Ft Riley avian point transects
description: |
  Ft Riley 2020 avian survey using covariates for bout-specific estimates.
author:
  - name: Rexstad 
    url: 
    affiliation: CREEM, University of St Andrews
    affiliation_url: https://www.creem.st-andrews.ac.uk/
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 1
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 90)
solution <- FALSE
```

Load the packages to be used

```{r packages}
library(Distance)
library(readxl)
library(knitr)
library(kableExtra)
```

Acquire data from the spreadsheet.  Note how `na=` argument is used.

```{r read}
spreadsheet.file <- "BBS_Observations_Master.xlsx"
birds <- as.data.frame(read_xlsx(path=spreadsheet.file, sheet="Observaitons", na="NA"))
```

Make adjustments to column names to conform to convention used in the `Distance` package.

```{r adjust}
names(birds)[7] <- "obsname"
birds$Effort <- birds$effort
birds$Region.Label <- birds$Region
birds$Sample.Label <- birds$pid
birds$Area <- 1
```

Specify units of measure.  With point transects there are no distances associated with `Effort`, that argument is set to `NULL`.

```{r conv}
birdunits <- convert_units("meter", NULL, "square kilometer")
```


There are also a couple of functions I will define here to be used later in our analysis.

```{r handy}
cvmtest <- function(dsobject) {
#  Small function that prints Cramer-von Mises test and associated P-value
  cvmw <- round(gof_ds(dsobject, plot=FALSE)$dsgof$CvM$W, 3)
  cvmp <- round(gof_ds(dsobject, plot=FALSE)$dsgof$CvM$p, 3)
  print(paste("CvM W=", cvmw, "P=", cvmp))
}

visitplot <- function(dsobject) {
# plots visit-specific detection functions
  plot(dsobject, pdf=TRUE, showpoints=FALSE, main="Detection function with visit as covariate")
  add_df_covar_line(dsobject, data=data.frame(bout=as.factor(1)), 
                    pdf=TRUE, lty=1, col="blue")
  add_df_covar_line(dsobject, data=data.frame(bout=as.factor(2)), 
                    pdf=TRUE, lty=1, col="green")
  add_df_covar_line(dsobject, data=data.frame(bout=as.factor(3)),
                    pdf=TRUE, lty=1, col="red")
  legend("topright", lty=1, legend=c("Visit 1", "Visit 2", "Visit 3"),
         col=c("blue", "green", "red"))
}
```

# Northern bobwhite

## Treating visit as stratum

```{r nobovisitstrat, eval=solution}
birds$Effort <- 1
mytruncation <- "10%"
nobo <- birds[(birds$species=="NOBO")  | is.na(birds$distance), ]
nobo.vstrat <- ds(data=nobo,
                transect="point", key="hr", formula=~as.factor(bout),
                truncation = mytruncation, convert.units = birdunits)
nobo.ests <- dht2(ddf=nobo.vstrat, flatfile=nobo,
                  strat_formula = ~as.factor(bout), convert_units = birdunits,
                  stratification = "replicate")
print(nobo.ests, report="density")
```

```{r noboplot, fig.cap="Northern bobwhite detection function with visit as a covariate", eval=solution}
cvmtest(nobo.vstrat)
visitplot(nobo.vstrat)
```

# Dickcissel

## Treating visit as stratum

Assuming detection functions for each visit share a common key function.

```{r dickvisitstrat, eval=solution}
dick <- birds[(birds$species=="DICK")  | is.na(birds$distance), ]
dick.vstrat <- ds(data=dick,
                transect="point", key="hr", formula=~as.factor(bout),
                truncation = mytruncation, convert.units = birdunits)
dick.ests <- dht2(ddf=dick.vstrat, flatfile=dick,
                  strat_formula = ~as.factor(bout), convert_units = birdunits,
                  stratification = "replicate")
print(dick.ests, report="density")
```

```{r visitplot, fig.cap="Dickcissel detection function with visit as a covariate", eval=solution}
cvmtest(dick.vstrat)
visitplot(dick.vstrat)
```

# Brown headed cowbird

## Visit as stratum covariate

Adjust effort for per-visit analysis

```{r bhcovisit, eval=solution}
bhco <- birds[(birds$species=="BHCO")  | is.na(birds$distance), ]
bhco.vstrat <- ds(data=bhco,
                transect="point", key="hr", formula=~as.factor(bout),
                truncation = mytruncation, convert.units = birdunits)
bhco.ests <- dht2(ddf=bhco.vstrat, flatfile=bhco,
                  strat_formula = ~as.factor(bout), convert_units = birdunits,
                  stratification = "replicate")
print(bhco.ests, report="density")
```

```{r visitplotbhco, fig.cap="Brown headed cowbird detection function with visit as a covariate", eval=solution}
cvmtest(bhco.vstrat)
visitplot(bhco.vstrat)
```

# Eastern meadowlark

## Visit as stratum covariate

Adjust effort for per-visit analysis

```{r eamevisit, eval=solution}
eame <- birds[(birds$species=="EAME")  | is.na(birds$distance), ]
eame.vstrat <- ds(data=eame,
                transect="point", key="hr", formula=~as.factor(bout),
                truncation = mytruncation, convert.units = birdunits)
eame.ests <- dht2(ddf=eame.vstrat, flatfile=eame,
                  strat_formula = ~as.factor(bout), convert_units = birdunits,
                  stratification = "replicate")
print(eame.ests, report="density")
```

```{r visitploteame, fig.cap="Eastern meadowlark detection function with visit as a covariate", eval=solution}
cvmtest(eame.vstrat)
visitplot(eame.vstrat)
```

# Bell's vireo
## Visit as stratum covariate

Adjust effort for per-visit analysis

```{r bevivisit, eval=solution}
bevi <- birds[(birds$species=="BEVI")  | is.na(birds$distance), ]
bevi.vstrat <- ds(data=bevi,
                transect="point", key="hr", formula=~as.factor(bout),
                truncation = mytruncation, convert.units = birdunits)
bevi.ests <- dht2(ddf=bevi.vstrat, flatfile=bevi,
                  strat_formula = ~as.factor(bout), convert_units = birdunits,
                  stratification = "replicate")
print(bevi.ests, report="density")
```

```{r visitplotbevi, fig.cap="Bells vireo detection function with visit as a covariate", eval=solution}
cvmtest(bevi.vstrat)
visitplot(bevi.vstrat)
```

# Henslow's sparrow

## Visit as stratum covariate

Adjust effort for per-visit analysis

```{r hespvisit, eval=solution}
hesp <- birds[(birds$species=="HESP")  | is.na(birds$distance), ]
hesp.vstrat <- ds(data=hesp,
                transect="point", key="hn", formula=~as.factor(bout),
                truncation = mytruncation, convert.units = birdunits)
hesp.ests <- dht2(ddf=hesp.vstrat, flatfile=hesp,
                  strat_formula = ~as.factor(bout), convert_units = birdunits,
                  stratification = "replicate")
print(hesp.ests, report="density")
```

```{r visitplothesp, fig.cap="Henslows sparrow detection function with visit as a covariate", eval=solution}
cvmtest(hesp.vstrat)
visitplot(hesp.vstrat)
```

The plot of bout-specific detection functions makes the estimation enterprise dubious.

```{r histos, eval=solution}
hesp <- birds[(birds$species=="HESP"), ]
h1 <- hesp[hesp$bout == 1, "distance"]
h2 <- hesp[hesp$bout == 2, "distance"]
h3 <- hesp[hesp$bout == 3, "distance"]
par(mfrow=c(1,3))
hist(h1, main="HESP, bout 1", nc=20)
hist(h2, main="HESP, bout 2", nc=20)
hist(h3, main="HESP, bout 3", nc=20)
par(mfrow=c(1,1))
```

# Addendum

What are the consequences of employing covariates in the detection function (beyond covariate for sampling session)?  Many other covariates were collected at the time of data collection, presumably because of their potential explanatory power in detection function modelling.  I will demonstrate the effect of a couple of covariates on the detection function of dickcissel.

```{r eda, echo=FALSE, layout="l-body-outset", fig.height=6, fig.cap="Exploratory data analysis of dickcissel detection function covariates."}
library(vioplot)
dick <- birds[(birds$species=="DICK")  | is.na(birds$distance), ]
par(mfrow=c(3,2))
vioplot(distance~obsname, data=dick)
vioplot(distance~scribe, data=dick)
plot(distance~wind, data=dick[dick$wind<10, ])  # some data errors for wind
plot(distance~start, data=dick)
plot(distance~temp, data=dick[dick$temp>10, ]) # really cold on 28June20
par(mfrow=c(1,1))
```

Nothing is evident in the exploratory plots; I suspect nothing will arise from covariate modelling.  To demonstrate the point, a barrage of single covariate models with above covariates (and hazard rate key), compared with pooled hazard rate without covariates:

```{r covarmods, echo=FALSE, error=FALSE}
mytruncation <- "10%"
dick <- dick[dick$wind<10,]
dickobs <- ds(dick, transect="point", key="hr", truncation = mytruncation, convert.units = birdunits,
              formula=~obsname)
dickscribe <- ds(dick, transect="point", key="hr", truncation = mytruncation, convert.units = birdunits,
                 formula=~scribe)
dickwind <- ds(dick, transect="point", key="hr", truncation = mytruncation, convert.units = birdunits,
               formula=~scale(wind))
dickstart <- ds(dick, transect="point", key="hr", truncation = mytruncation, convert.units = birdunits,
                formula=~scale(start))
dicktemp <- ds(dick, transect="point", key="hr", truncation = mytruncation, convert.units = birdunits,
               formula=~scale(temp))
dicknothing <- ds(dick, transect="point", key="hr", truncation = mytruncation, convert.units = birdunits)
kable(summarize_ds_models(dickobs, dickscribe, dickwind, dickstart, dicktemp, dicknothing)[, 2:7], row.names = FALSE, 
      digits=c(0,0,2,5,2,5), caption="Model selection for covariate models with Dickcissel 2020")
```

Look not only at the $\Delta$AIC but also at the differences in $\hat{P_a}$.  These differences translate into differences in $\hat{D}$

```{r outests, echo=FALSE}
outests <- rbind(dickobs$dht$individuals$D[,2:6],
                 dickscribe$dht$individuals$D[,2:6],
                 dickwind$dht$individuals$D[,2:6],
                 dickstart$dht$individuals$D[,2:6],
                 dicktemp$dht$individuals$D[,2:6],
                 dicknothing$dht$individuals$D[,2:6])
row.names(outests) <- c("Observer", "Scribe", "Wind", "Start", "Temp", "Nothing")
kable(outests, digits=3, caption="Density estimates for Dickcissel from various covariate models")
```